---
title: "Efficient Programming in R"
subtitle: "BioMarin R-Meetup: tips for improving your code"  
author: 
  - "Martin Frigaard"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(reprex)
library(xaringan)
library(pagedown)
fs::dir_create("img")
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center", 
                      dpi = 320, fig.height = 4, fig.path = "img/")

```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
xaringanthemer::style_mono_light(base_color = "cornflowerblue")
# xaringan::inf_mr() # for instant knitting and reviewing
```

class: center, middle

# Welcome to efficient R programming!

---
class: left, top

# Objectives

### 1. Define efficiency/efficient programming in R  

### 2. Practices for improving programming efficiency  

### 3. Packages to improve your efficiency  


---
class: left, top

# What is the 'efficient' programming in R?

--

### *Efficient programming, defined:*

--

> ### ***"...the amount of useful work a person (not a computer) can do per unit time"***

--

### - *Efficient R programming, defined:*

--

> #### - ***...efficient R programming can be interpreted as finding a solution that is fast enough in terms of computational efficiency but as fast as possible in terms of programmer efficiency.***

Source: [Efficient R Programming, 2020](https://csgillespie.github.io/efficientR/introduction.html)

---
class: left, top









