---
title: "BioMarin Meetup: reshaping, transforming, and joining data in R (exercises)"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: tango
    theme: lumen
    df_print: paged
    
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(skimr)
# create data folder
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      tidy = FALSE)
```

# Objectives 

This document outlines some common data reshaping tasks with `tidyr`. The slides for this presentation are [here](https://mjfrigaard.github.io/data-transformations-joins/Index.html#1)

There is also an accompanying RStudio.Cloud project: 

Read more about `dplyr` [here](https://dplyr.tidyverse.org/index.html) and [`tidyr`](https://tidyr.tidyverse.org/) on the tidyverse website, or in the [Data Transformation](https://r4ds.had.co.nz/transform.html) chapter of R for Data Science. 

## Load packages 

`tidyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. 

```{r packages, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```


## Manging data {.tabset}

We're going to use some of the `fs` package functions for managing our files and data. 

### Folder trees and paths

Check your files and folders with `dir_tree` or `dir_ls`. 

```{r dir_tree}
fs::dir_tree(".", recurse = FALSE)
```

We can also use `dir_ls` to get the paths to print to the screen so we can easily import. 

```{r dir_ls}
fs::dir_ls(".", recurse = FALSE)
```

### Metadata 

`fs` also gives us the metadata of any folder we have access to. This can come in handy if we're looking for a specific version of a file, or a file that was created at a specific date. 

```{r fs-fir-info}
fs::dir_info("data", all = TRUE) %>% 
    dplyr::arrange(desc(modification_time))
```

For example, lets look for the largest `file` in the `data` folder. 

```{r largest-file}
fs::dir_info("data", all = TRUE) %>% 
    filter(type == "file") %>% 
    arrange(desc(size)) %>% 
    select(path, size) %>% 
    slice(1)
```

### Import multiple files 

Lets import all the `Lab` data files in `data` using `fs` and `purrr`. 

```{r import-labs}
fs::dir_ls(path = "data", regexp = "Lab") %>% 
    purrr::map_dfr(read_csv, .id = "source") -> AllLabs
AllLabs
```

***

## `separate()`  {.tabset}

This allows us to split columns apart into multiple variables. 

### `separate` ex 1

Take a look at `SmallLabData` below (we've used some `dplyr` and the `janitor::remove_empty()` function to get `SmallLabData` from `AllLabs`.

```{r SmallLabData}
AllLabs %>% 
    filter(source == "data/SmallLabData.csv") %>% 
    select(-source) %>% 
    janitor::remove_empty() -> SmallLabData
SmallLabData
```


Separate `ref` into `ref_value` and `ref_units`

```{r separate-ex, eval=FALSE}
SmallLabData %>% 
    separate(col = ___, into = c("_______", "_______"), sep = " ")
```


### `separate` sol 1

```{r separate-sol}
SmallLabData %>% 
    separate(col = ref, into = c("ref_value", "ref_units"), sep = " ")
```

***

## `separate_rows()`  {.tabset}

`separate_rows()` is similar to `separate()`, but drops the values down the rows. This is commonly used when we have more than one value in each column. 

Take a look at `LabData` below. 

```{r LabData}
AllLabs %>% 
    filter(source == "data/LabData.csv") %>% 
    select(-source) %>% 
    janitor::remove_empty() -> LabData

 LabData
```

### `separate_rows` ex 1

Put all the `value`s into their own cell, and make sure the new variable is numeric. 

```{r separate_rows-ex, eval=FALSE}
LabData %>% 
  separate_rows(_____, sep = _____, convert = TRUE)
```


### `separate_rows` sol 1

Note the `convert` argument to make `value` numeric. 

```{r separate_rows-solution}
LabData %>% 
  separate_rows(value, sep = ", ", convert = TRUE)
```

***

## `unite()` {.tabset} 

Just like `separate`/`separate_rows` allows us to split up columns, `unite()` lets us stick them back together!

Take a look at `LabProc`.

```{r LabProc}
AllLabs %>% 
    filter(source == "data/LabProc.csv") %>% 
    select(-source) %>% 
    janitor::remove_empty() -> LabProc
LabProc
```

### `unite` ex 1

Separate the labs from each other, and drop them down the rows. 

```{r unite-separate_rows, eval=FALSE}
LabProc %>% 
    separate_rows(lab_proc, sep = "___")
```

### `unite` ex 2

Put `lab`, `value`, and `units` in to their own columns.

```{r unite-separate_rows-separate, eval=FALSE}
LabProc %>% 
    separate_rows(lab_proc, sep = "___") %>% 
    separate(col = lab_proc,into = c('__', '_____', '_____'), 
           sep = " ", convert = ____)
```

### `unite` ex 3

Unite `lab`, `value` and `units` into `lab_name` (don't remove the previous columns with `remove = FALSE`).

```{r unite, eval=FALSE}
LabProc %>% 
    separate_rows(lab_proc, sep = "___") %>% 
    separate(col = lab_proc,into = c('__', '_____', '_____'), 
           sep = " ", convert = ____)
    unite(col = "_______", lab:units, sep = " ", remove = _____)
```

### `unite` sol 1

Separate the labs from each other, and drop them down the rows.

```{r unite-solution-1}
LabProc %>% 
     separate_rows(lab_proc, sep = "; ") 
```

### `unite` sol 2

Put `lab`, `value`, and `units` in to their own columns.

```{r unite-solution-2}
LabProc %>% 
     separate_rows(lab_proc, sep = "; ") %>% 
     separate(col = lab_proc,into = c('lab', 'value', 'units'), 
            sep = " ", convert = TRUE) 
```

### `unite` sol 3

Unite `lab`, `value` and `units` into `lab_name` (don't remove the previous columns with `remove = FALSE`).

```{r unite-solution-3}
LabProc %>% 
     separate_rows(lab_proc, sep = "; ") %>% 
     separate(col = lab_proc,into = c('lab', 'value', 'units'), 
            sep = " ", convert = TRUE) %>% 
     unite(col = "lab_name", lab:units, sep = " ", remove = FALSE)
```

***

## `unnest()` use case

`unnest()` deals with nested data, which can be complicated to explain, so we'll use an example. Run the code below to download data from the [Anti-Doping Agency Sanction database](https://www.usada.org/news/sanctions/). 

```{r 01-create-usada-data}
source("code/01-create-usada-data.R")
```

These data have two dates (`original` and `updated`) in a single column (`sanction_announced`). **We only want `updated` dates in their own columns.**

```{r UsadaBadDates}
UsadaBadDates
```

### `str_split` 

Use the `stringr` package's `str_split` function to split the bad dates on the `"updated"` pattern. 

```{r str_split}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # only return what we're interested in 
    dplyr::select(athlete, sanction_announced, sanction_dates)
```

### Nested data

This creates a nested column (`sanction_dates`), which we can deal with by `unnest()`ing it. Let's take a better look with `listviewer` (but only look at `athlete` and `sanction_dates`).

```{r listviewer}
library(listviewer)
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # only return what we're interested in 
    dplyr::select(athlete, sanction_dates) %>% 
    listviewer::jsonedit(mode = "view")
```


### `unnest`

After unnesting, we can see the dates have been repeated down the rows (like `separate_rows()`).

```{r unnest}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    # check this to see what is happening 
    dplyr::select(athlete, sanction_announced, sanction_dates)
```

### `filter` 

We can see all of the updated dates are on a new line, and the original dates still contain the text. Now I can filter out the rows with an original date (because I am only interested in the updated dates).

```{r filter}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # 2) convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    
    # 3) remove the "original" dates 
        dplyr::filter(!str_detect(string = sanction_dates, 
                                  pattern = "original")) %>% 
    # check this to see what is happening 
    dplyr::select(athlete, sanction_announced, sanction_dates)
```

### `str_remove_all`

Now I can clean up the updated `sanction_dates` and format it. 

```{r str_remove_all}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # 2) convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    
    # 3) remove the "original" dates 
        dplyr::filter(!str_detect(string = sanction_dates, 
                                  pattern = "original")) %>% 
    # 4) remove the colon from sanction_dates
        dplyr::mutate(sanction_dates = stringr::str_remove_all(
                                                      string = sanction_dates,
                                                      pattern = ":"),
                      # 5) remove any whitespace
                      sanction_dates = stringr::str_trim(sanction_dates),
                      # 6) format as date
                      sanction_dates = lubridate::mdy(sanction_dates)) %>%
    # check this to see what is happening 
    dplyr::select(athlete, sanction_announced, sanction_dates)
```


### `ggplot2`

***How far can this pipe go??***

Well, when the data are arranged in the way we want, we can test it with a visualization to check the number of sanctions per sport with bad dates. 

```{r visualize}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # 2) convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    
    # 3) remove the "original" dates 
        dplyr::filter(!str_detect(string = sanction_dates, 
                                  pattern = "original")) %>% 
    # 4) remove the colon from sanction_dates
        dplyr::mutate(sanction_dates = stringr::str_remove_all(
                                                      string = sanction_dates,
                                                      pattern = ":"),
                      # 5) remove any whitespace
                      sanction_dates = stringr::str_trim(sanction_dates),
                      # 6) format as date
                      sanction_dates = lubridate::mdy(sanction_dates)) %>% 
    # count by sport
    dplyr::count(sport, sort = TRUE) %>% 
    # rename n 
    rename(`Sport Count` = n,
            Sport = sport) %>% 
    ggplot2::ggplot(aes(x = `Sport Count`, 
                        y = forcats::fct_reorder(as.factor(Sport), `Sport Count`), 
                        fill = as.character(`Sport Count`))) + 
    geom_col(show.legend = FALSE) + 
    labs(x = "Number of Sanctions", y = "Sport", 
         title = "Number of Sanctions with Bad Dates",
         caption = "source: https://www.usada.org/news/sanctions/")
```


