---
title: "BioMarin Data Manipulation Training: An Introduction to Wrangling Data"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    highlight: tango
    theme: lumen
    df_print: paged
    
params: 
  data_file: !r file.path("https://bit.ly/mini-strwrs")

always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(ipcwswitch)
library(skimr)
# create data folder
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      tidy = FALSE,
                      size = "Huge")

original_starwars <- read_csv("https://bit.ly/mini-strwrs")
```

# Objectives 

This document outlines some common data manipulation tasks with `dplyr`. The slides for this presentation are [here](https://mjfrigaard.github.io/data-manip-intro/Index.html#1). There is also an accompanying RStudio.Cloud project [here.](https://rstudio.cloud/project/1918833) 



Read more about `dplyr` [here](https://dplyr.tidyverse.org/index.html) on the tidyverse website, or in the [Data Transformation](https://r4ds.had.co.nz/transform.html) chapter of R for Data Science. 

## Load packages 

`dplyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. 

```{r packages, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```


## Import data {.tabset}

We have the path to the `original_starwars` data stored in our `params`, but we will also go over how to build this dataset from `dplyr::starwars`. 

### Import using url

Below we import the `original_starwars` dataset from the slides using the url. This is similar to providing a local file path (`data/original-starwars.csv`).

```{r import-url}
read_csv("https://bit.ly/mini-strwrs")
```

### Import using `params`

We have the `params` list from our YAML header, which we can also use to import the data. 

```{.yaml}
params: 
  data_file: !r file.path("https://bit.ly/mini-strwrs")
```

```{r import-params}
readr::read_csv(params$data_file)
```

### Create from `dplyr`

We can also manually create `original_starwars` using the code below (which we cover in this training). 

```{r manually-create-original_starwars}
dplyr::starwars %>% 
    dplyr::select(name, height, mass, hair_color, species, homeworld) %>% 
    filter(name %in% c("Luke Skywalker", "C-3PO", "R2-D2", 
                       "Leia Organa", "Chewbacca", "Han Solo"))
```

## Selecting columns {.tabset}

These are exercises to try on your own using the `select()` function.

### Select exercise 1

Alter the code below to select just the `name` and `homeoworld` column:

```{r select-ex-01, eval=FALSE}
select(original_starwars, name, species, homeworld)
```


### Select solution 1

```{r select-sol-01}
select(original_starwars, name, homeworld)
```


### Select exercise 2

Select only the columns starting with the letter `h`. 

```{r select-ex-02, eval=FALSE}
select(original_starwars, starts_with("_"))
```


### Select solution 2

```{r select-sol-02}
select(original_starwars, starts_with("h"))
```

## Filtering rows {.tabset}

These are some additional exercises for `filter()`ing data. 

### Filtering exercise 1

Change the code below so `original_starwars` only includes the droids. 

```{r filter-ex-01, eval=FALSE}
filter(original_starwars, species == "____")
```


### Filtering solution 1

```{r filter-sol-01}
filter(original_starwars, species == "Droid")
```

### Filtering exercise 2

Change the code below so `original_starwars` only includes data from the `homeworld`s of `Tatooine` and `Alderaan`

```{r filter-ex-02, eval=FALSE}
filter(original_starwars, 
       homeworld %in% c("________", "________"))
```

### Filtering solution 2

```{r filter-sol-02}
filter(original_starwars, 
       homeworld %in% c("Tatooine", "Alderaan"))
```


## Arranging data {.tabset}

`arrange()` sorts variables on their content, numeric or character. 

### Arrange exercise 1 

Sort `original_starwars` according to the `hair_color`. 

```{r arrange-ex-01, eval=FALSE}
arrange(original_starwars, "____ _____")
```


### Arrange solution 1 

Note that the missing values are sorted to the bottom. 

```{r arrange-sol-01}
arrange(original_starwars, hair_color)
```


### Arrange exercise 2

Sort `original_starwars` by `height` and `mass`, descending. 

```{r arrange-ex-02, eval=FALSE}
arrange(original_starwars, desc(______, ____))
```

### Arrange solution 2

```{r arrange-sol-02}
arrange(original_starwars, desc(height, mass))
```

Including two variables is helpful if some of the values 'tie'. 

## Mutate columns {.tabset}

`mutate()` can create new columns, or change existing columns.  

### Mutate exercise 1 

Alter the code below to create a `bmi` column for `starwars` characters in `original_starwars`.

```{r mutate-ex-01, eval=FALSE}
mutate(original_starwars, 
       bmi = ____ / ((______ / 100) ^ 2))
```

### Mutate solution 1

Note the use of parentheses here. 

```{r mutate-sol-01}
mutate(original_starwars, 
       bmi = mass / ((height / 100) ^ 2))
```

### Mutate exercise 2

Round the new `bmi` variable to 1 digit. 

```{r mutate-ex-02, eval=FALSE}
mutate(original_starwars, 
       bmi = mass / ((height / 100) ^ 2),
       bmi = _____(___, digits = _))
```

### Mutate solution 2

```{r mutate-sol-02}
mutate(original_starwars, 
       bmi = mass / ((height / 100) ^ 2),
       bmi = round(bmi, digits = 1))
```

## Create function pipelines with %>% {.tabset}

The pipe (`%>%`) takes an object that comes before it, and it drops the object into the function that comes after it.

### Pipe exercise 1 

Rewrite the code below to use the pipe 

```{r pipe-ex-01, eval=FALSE}
select(filter(
          original_starwars, mass < ___), 
              ____, hair_color, _______, homeworld)
```

### Pipe solutons 1

This actually wouldn't matter what order it was in--both would return the same result. 

```{r pipe-sol-01}
original_starwars %>% 
  filter(mass < 100) %>% 
  select(name, hair_color, species, homeworld)
```


### Pipe exercise 2

Perform the following operations *without* using the pipe. 

1. create object `x` with three values, (`3`, `7`, `12`)    
2. get the `mean()` of `x`, and store it in `mean_x`   
3. take the squared root of `mean_x`  

```{r pipe-ex-02, eval=FALSE}
# 1) 
_ <- c(_, _, __)
# 2) 
mean_x <- ____(x)
# 3 
sqrt(______)
```

### Pipe soluton 2

This returns a vector, not a `tibble`.

```{r pipe-sol-02}
# 1) 
x <- c(3, 7, 12)
# 2) 
mean_x <- mean(x)
# 3 
sqrt(mean_x)
```

### Pipe exercise 3

Perform the following operations *with* the pipe. 

1. create object `x` with three values, (`3`, `7`, `12`)    
2. get the `mean()` of `x`, and store it in `mean_x`   
3. take the squared root of `mean_x`

```{r pipe-ex-03, eval=FALSE}
c(_, _, __) %>% 
  ____() %>% 
  ____()
```


### Pipe solutons 3

Note that we can create a pipeline without even creating an object. 

```{r pipe-sol-03}
c(3, 7, 12) %>% 
  mean() %>% 
  sqrt()
```